Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-07-23 manuscript/manuscript.tex
Add FWF Funding Information
small changes, mostly to grammar, but some effect the meaning of sentences

GIST has been implemented on the GPU, substantially improving its performance for large grids \cite{Kraml2019-gigist}.
Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}. 
Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.

@@

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}
Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:

\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
\end{lstlisting}
\begin{minipage}{\linewidth} %TODO: maybe find a better way to keep listing on same page?
Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. Any preprocessing steps need to match between the \inlinecode{bounds} command and the \inlinecode{gist} calculation. A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
~
\begin{lstlisting}[style=bash]
python findcentroid.py -i structure.pdb
\end{lstlisting}
\end{minipage}
\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience. The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

>>>>

GIST has been implemented on the GPU, substantially improving its performance for large grids \cite{Kraml2019-gigist}.
Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}. 
Other extensions of GIST allow the inclusion of rigid solvents other than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.

@@

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While all MD simulations require a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

Furthermore, a recent Python version (3.6 or newer) should be installed, along with the following packages:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
The visualization of three-dimensional contributions to solvation free energy requires molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were presented in a previous tutorial on GIST \cite{Ramsey2016}.

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}
Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in a binding pocket, or an entire protein. 
Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:

\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
\end{lstlisting}
\begin{minipage}{\linewidth} %TODO: maybe find a better way to keep listing on same page?
Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. Any preprocessing steps need to match between the \inlinecode{bounds} command and the \inlinecode{gist} calculation. A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
~
\begin{lstlisting}[style=bash]
python findcentroid.py -i structure.pdb
\end{lstlisting}
\end{minipage}
\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience and reproducibility. The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
